---
title: "Formal analysis -- 1. Rarefaction analysis"
subtitle: ""
date: "`r Sys.Date()`"
author: Yue (Gary) Zhang, PhD
output:
  rmdformats::readthedown:
    self_contained: true
    thumbnails: true
    lightbox: true
    gallery: false
    highlight: tango
    code_folding: hide
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = T, message = T,echo = T) 
```


```{r Initiate, results='hide'}
## initiate ---------------------------------------------------------
packages_to_load = c('dplyr','ggplot2','vegan','ape','stringr','usedist','ggpubr','corrplot',
                     'cowplot','GGally','scales','metagMisc','ggsci','rstatix','wesanderson', 'readxl','tidyr',
                     'phyloseq','DT','plotly','ranacapa','ggrepel')
invisible(suppressPackageStartupMessages({
    lapply(packages_to_load, library, character.only = TRUE)
}))

#generating color palettes
col1 = wes_palette("Rushmore1")
col2 = wes_palette("GrandBudapest2")
col3 = wes_palette('Darjeeling2')
col21 <- rev(c("tomato1","darkblue","turquoise1","lightblue","darkred","mediumblue","purple","bisque",
               "greenyellow","yellow","violetred2","darkgreen","darkgoldenrod1","deeppink3","cadetblue4",
               "orchid2","seagreen3","purple4","dodgerblue2","red","gray27"))

set.seed(666)
```

```{r functions,eval=TRUE}
source('utils_functions.R')
```

```{r variables}
indir = "../data/R_input"
outdir = "../data/Formal_analysis_files"

if (file.exists(outdir)){
  unlink(outdir,recursive = TRUE)
  dir.create(outdir)
} else {
  dir.create(outdir)
}


```

```{r import}
metadata.raw = read.csv(file.path(indir,'metadata.unparsed.withHC.txt'),sep = '\t')
metadata.additional = read.csv(file.path(indir,'FINAL_8.28.24_NIH_BAL_combined_taxa_table.SORTED_COVID.csv'))
otu.raw = read.csv(file.path(indir,'nih_bal.shared.txt'),sep = '\t')
tree= read.tree(file.path(indir,'nih_bal_mothur.rep.phylip.tre')) # Dickson's group-generated phylogenetic tree
taxa.raw = read.csv(file.path(indir,'nih_bal.cons.taxonomy.txt'),sep = '\t')
```


```{r parse}
# metadata -----------
metadata.raw = merge(metadata.raw,
                     select(metadata.additional, c(uniqID, ID, Date, Stage)),
                     by = 'uniqID',all.x = T)

## remove the duplicate measurements of two controls (specimen_BAL_09e6hL and 03C; their counter parts have a higher number of reads, which are kept)
metadata.raw = subset(metadata.raw,!uniqID %in% c('09e6hL','03C'))
## set a consistent rowname to avoid confusions downstream
metadata.raw$sample.id.new = paste0('s',seq_len(nrow(metadata.raw)))
rownames(metadata.raw) <- metadata.raw$sample.id.new
sample_table = sample_data(metadata.raw) # this leads to automatic renaming the samples, so we have to rename the feature table correspondingly
metadata = as.data.frame(as.matrix(sample_table))
```


```{r otu phylotree taxa parsing}
# OTU table --------
names(otu.raw)[1] <- 'sample.id'
# sanity check 
otu = subset(otu.raw, sample.id %in% metadata$sample.id)
stopifnot(length(intersect(metadata$sample.id, otu$sample.id)) == nrow(metadata))
otu <- setrowname(otu)

otu <- otu %>%
  as.matrix %>%
  t 

colnames(otu) = rownames(sample_table)[match(colnames(otu),sample_table$sample.id)]
otu= otu[,rownames(sample_table)]
feature_table = otu_table(as.matrix(otu),taxa_are_rows = T)

# phylogenetic tree -----------
# Dickson group generated the tree
# remove long sufix in tree tip labels
tree$tip.label = sub('_.*','',tree$tip.label)

# taxa table ------------
taxa <- taxa.raw %>% 
  separate(Taxonomy, c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Species"), sep = ';') %>%
  mutate_all(~ ifelse(. == "", "Unclassified", .)) %>%               # Replacing empty cells
  mutate_all(~ gsub("\\s*\\(.*?\\)", "", .))                  # Removing everything in parentheses
  #mutate_all(~ ifelse(grepl("unclassified", ., ignore.case = TRUE), "Unclassified", .)) # Ensuring "unclassified" is replaced

rownames(taxa) = taxa$OTU
taxa = taxa[,-c(1:2)]
taxa_table = tax_table(as.matrix(taxa))

# construct phyloseq objects -------------------
p.all = phyloseq(feature_table,tree,taxa_table,sample_table)
```

# Remove a few duplicated samples

```{r count reads}
p.all = count_total_reads(p.all,'raw')
metatable = as.data.frame(as.matrix(sample_data(p.all)))
metatable$read_count_raw = as.numeric(metatable$read_count_raw)
ggboxplot(metatable,x='specimen_type',y='read_count_raw')+
  scale_y_log10()
```

There are three duplicated samples: ID=189 at stage R, ID=117 at stage C, and ID=223 at stage A. sample of 117_C (s153) with 3 reads is removed here (it will be removed during rarefaction anyway). For the other two pairs, we merge them by adding OTU read count up.

```{r dedup}
sample_data(p.all)[['ID_Stage']] = paste0(sample_data(p.all)[['ID']],'_',sample_data(p.all)[['Stage']])
metatable = as.data.frame(as.matrix(sample_data(p.all)))

dc = as.data.frame(table(metatable$ID_Stage))
dc = subset(dc,Freq>1 & Var1 != 'NA_NA')
metatable = subset(metatable, ID_Stage %in% dc$Var1, select = c(ID_Stage,read_count_raw,sample.id.new))
create_dt(metatable, "Duplicates")

# separate the non-redundant samples with the duplicated ones (keep s154)
p.all.temp = prune_samples(c(setdiff(sample_names(p.all),metatable$sample.id.new),"s154"),p.all)

p.all.189R = merge_duplicates(rownames(subset(metatable,ID_Stage == '189_R')),p.all)
p.all.223A = merge_duplicates(rownames(subset(metatable,ID_Stage == '223_A')),p.all)

p.all.dedup = merge_phyloseq(p.all.189R,p.all.223A)
sample_data(p.all.dedup)$sample.id.new = rownames(sample_data(p.all.dedup))
metatable = as.data.frame(as.matrix(sample_data(p.all.dedup)))
p.all = merge_phyloseq(p.all.temp,p.all.dedup)
metatable = as.data.frame(as.matrix(sample_data(p.all)))

```



```{r save p all}
saveRDS(p.all,file = file.path(outdir,'phyloseqObj.all_samples.deduplicated.unrarefied.rds'))
```

# Rarefaction

## Sample counts in the potential range of rarefaction depths (1000-5000)


```{r sample combination count}
# select samples
p.1 = subset_samples(p.all,specimen_control == 'specimen')
metatable <- p.1 %>%
  sample_data %>%
    as.matrix %>%
    as.data.frame
metatable$read_count_raw = as.numeric(metatable$read_count_raw)
```

```{r rr sample count}
res.all = NULL
for (depth in c(5000,4000,3000,2000,1000,500,0)){
  print(depth)
  df = subset(metatable,read_count_raw>depth)
  res = calculate_stage_combinations(df,depth)
  #print(res)
  if (!is.null(res.all)){
    res.all = merge(res,res.all,by = 'Stage_Combination',all=T)
  } else {res.all = res}
}

create_dt(res.all,'Sample counts at different rarefaction depth')

```

## Manual rarefaction

```{r rarefy, fig.width=16,fig.height=14,eval=TRUE}

p.1 = count_total_reads(p.1,'before_rarefaction')
create_dt(as.data.frame(sample_data(p.1)),'Metadata')


# Define rarefaction depths and calculate alpha diversity metrics
depths <- seq(500, 20000, by = 200)

rarefying_list <- lapply(depths, function(d) {
  richness <- p.1 %>%
    rarefy_even_depth(., sample.size = d, rngseed = T, replace = F, trimOTUs = TRUE, verbose = F) %>%
    estimate_richness %>%
    select(c(Observed,Chao1,Shannon,InvSimpson)) %>%
    mutate(Rarefying_depth = d,
           Sample.id = rownames(.))
  #print(paste0('rarefying finished at depth ',d))
  return(richness)
  })
# merge list to a overarching dataframe
rarefaction = bind_rows(rarefying_list)
saveRDS(rarefaction,file.path(outdir,'rarefaction.RDS'))
```

```{r manual rarefy plot,fig.width=20,fig.height=14,eval=FALSE}
# Plot
#rarefaction_long = readRDS(file.path(outdir,'rarefaction.RDS'))
# Reshape the data from wide to long format
rarefaction_long <- rarefaction %>%
  pivot_longer(cols = c(Observed, Chao1, Shannon, InvSimpson), # Columns to stack
               names_to = "Metric",                           # New column for metric names
               values_to = "Value")     
p1 = ggplot(rarefaction_long, aes(x = Rarefying_depth, y = Value, color = Sample.id)) +
  geom_line() +
  labs(title = "Manual Rarefaction Curve", x = "Sampling Depth", y = "Metric Value") +
  theme(legend.position = "none") +
  facet_wrap(~ Metric, scales = "free_y")
ggplotly(p1)

```

```{r rarefy plots2}
rar.sha = ggplot(rarefaction, aes(x = Rarefying_depth, y = Shannon, color = Sample.id)) +
  geom_line() +
  labs(title = "Manual Rarefaction Curve -- Shannon Index", x = "Sampling Depth", y = "Shannon Index") +
  theme(legend.position = "none") 
rar.obs = ggplot(rarefaction, aes(x = Rarefying_depth, y = Observed, color = Sample.id)) +
  geom_line() +
  labs(title = "Manual Rarefaction Curve -- Observed OTUs", x = "Sampling Depth", y = "Observed OTUs") +
  theme(legend.position = "none") 
rar.InS = ggplot(rarefaction, aes(x = Rarefying_depth, y = InvSimpson, color = Sample.id)) +
  geom_line() +
  labs(title = "Manual Rarefaction Curve -- InvSimpson", x = "Sampling Depth", y = "InvSimpson") +
  theme(legend.position = "none") 

ggplotly(rar.obs)
ggplotly(rar.sha)
ggplotly(rar.InS)
```



## ggrare automatic rarefaction for benchmarking

```{r ggrare,results='hide'}
p = ggrare(p.1, step = 200, color = 'Stage', label = NULL, se = TRUE)
```

### Interactive plots

```{r ggrare plot,results='asis'}
cat('\n\n')
cat('#### ggrare full rarefaction curves')
cat('\n\n')
ggplotly(p)
cat('\n\n')
cat('#### ggrare rarefaction curves capped at 10,000 on Xaxis')
cat('\n\n')
p1 = p + xlim(0,10000)
ggplotly(p1)
```

# Session information and R image

```{r session info,eval=T}
sessionInfo()
save.image(file = file.path(outdir,"1.Rarefaction_analysis.RData"))
saveRDS(p.1,file = file.path(outdir,'phyloseqObj.specimen.deduplicated.unrarefied.rds'))

```