---
title: "Formal analysis -- 1.b.Metadata_rarefaction_diversityMetrics_calc"
subtitle: ""
date: "`r Sys.Date()`"
author: Yue (Gary) Zhang, PhD
output:
  rmdformats::readthedown:
    self_contained: true
    thumbnails: true
    lightbox: true
    gallery: false
    highlight: tango
    code_folding: hide
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = T, message = T,echo = T) 
```


```{r Initiate, results='hide'}
## initiate ---------------------------------------------------------
packages_to_load = c('dplyr','ggplot2','vegan','ape','stringr','usedist','ggpubr','corrplot',
                     'cowplot','GGally','scales','metagMisc','ggsci','rstatix','wesanderson', 'readxl','tidyr',
                     'phyloseq','DT','plotly','ranacapa','ggrepel','lmerTest')
invisible(suppressPackageStartupMessages({
    lapply(packages_to_load, library, character.only = TRUE)
}))

#generating color palettes
col1 = wes_palette("Rushmore1")
col2 = wes_palette("GrandBudapest2")
col3 = wes_palette('Darjeeling2')
col21 <- rev(c("tomato1","darkblue","turquoise1","lightblue","darkred","mediumblue","purple","bisque",
               "greenyellow","yellow","violetred2","darkgreen","darkgoldenrod1","deeppink3","cadetblue4",
               "orchid2","seagreen3","purple4","dodgerblue2","red","gray27"))

set.seed(666)
```

```{r functions,eval=TRUE}
source('utils_functions.R')
source('auxiliaries.R')
```

```{r variables}
indir = "../data/R_input"
outdir = "../data/Formal_analysis_files2"

# if (file.exists(outdir)){
#   unlink(outdir,recursive = TRUE)
#   dir.create(outdir)
# } else {
#   dir.create(outdir)
# }

# phyloseqObj.specimen.deduplicated.unrarefied.rds AND phyloseqObj.all_samples.deduplicated.unrarefied.rds are copied from Formal_analysis_files manually
p.1 = readRDS(file.path(outdir,'phyloseqObj.specimen.deduplicated.unrarefied.rds'))
```

# Add phenotype information

__Here we add the demographic information (V2_Demographics_August_2024_amended.csv) into the phyloseq object.__

For disease severity:

4= patients with mild disease/ambulatory ---> converted to normoxic

5, 6, and 7= moderate/severe disease ---> converted to hypoxic

Note that "specimen_BAL_09e6hL", "unknown_unknown_03C", and "specimen_BAL_17e24hL" are not in p.1r. The first two are removed as duplicated measurements from the beginning, and the third one only has about 200 reads so removed after rarefaction.

__Update the omicron status__

Using January 2022 as a cutoff (patients with symptom onset date before this date is pre-omicron).

```{r add metadata 1 }
metatable <- p.1 %>%
  sample_data %>%
  as.matrix %>%
  as.data.frame %>% 
  dplyr::select(c('sample.id','ID'))

demo = read.csv(file.path(indir,'V2_Demographics_August_2024_amended.csv'))
demo$disease_severity = ifelse(demo$Highest.ordinal.scale_Rec <= 4, 'Normoxic', 'Hypoxic')

demo = merge(metatable,demo,by= 'ID')
demoHC = read.csv(file.path(indir,'Demographics_Microbiome_HC.csv'))
demo= bind_rows(demo,demoHC)

med = read.csv(file.path(indir,'antibiotic_antiviral.usage.csv'))
med$ID = as.character(med$ID)

bal.clu <- read.csv('../data/R_input/olinkCluster.csv') %>% select(ID_Stage,BAL_Protein_Cluster)

sample_data(p.1)[['gender']] = demo$GENDER[match(sample_data(p.1)[['sample.id']],demo$sample.id)]
sample_data(p.1)[['gender']] = paste0('gender_', sample_data(p.1)[['gender']])

sample_data(p.1)[['race']] = demo$Race[match(sample_data(p.1)[['sample.id']],demo$sample.id)]
sample_data(p.1)[['race']] = paste0('race_',sample_data(p.1)[['race']])

sample_data(p.1)[['age']] = demo$Age[match(sample_data(p.1)[['sample.id']],demo$sample.id)]

sample_data(p.1)[['vaccination']] = demo$Vaccination[match(sample_data(p.1)[['sample.id']],demo$sample.id)]
sample_data(p.1)[['vaccination']] = ifelse(sample_data(p.1)[['vaccination']] == '0', 'Unvaccinated','Vaccinated')

sample_data(p.1)[['disease_severity']] = demo$disease_severity[match(sample_data(p.1)[['sample.id']],demo$sample.id)]

sample_data(p.1)[['antibiotics_antiviral_usage']] = med$antibiotics_antiviral_usage[match(sample_data(p.1)[['ID']],med$ID)]
sample_data(p.1)[['antibiotics_antiviral_usage']][is.na(sample_data(p.1)[['antibiotics_antiviral_usage']])] <- 'Not used'
sample_data(p.1)[['treatment']] = ifelse(sample_data(p.1)[['antibiotics_antiviral_usage']] == 'Not used','Untreated','Treated')

sample_data(p.1)[['BAL_Protein_Cluster']] = bal.clu$BAL_Protein_Cluster[match(sample_data(p.1)[['ID_Stage']],bal.clu$ID_Stage)]
sample_data(p.1)[['BAL_Protein_Cluster']][is.na(sample_data(p.1)[['BAL_Protein_Cluster']])] <- 'Not included'


sample_data(p.1)[['gender']] = as.factor(sample_data(p.1)[['gender']])
sample_data(p.1)[['race']]  = as.factor(sample_data(p.1)[['race']] )
sample_data(p.1)[['vaccination']] = as.factor(sample_data(p.1)[['vaccination']])
sample_data(p.1)[['age']] = as.numeric(sample_data(p.1)[['age']])
sample_data(p.1)[['disease_severity']] = as.factor(sample_data(p.1)[['disease_severity']])
sample_data(p.1)[['antibiotics_antiviral_usage']]  = as.factor(sample_data(p.1)[['antibiotics_antiviral_usage']] )
sample_data(p.1)[['treatment']] = as.factor(sample_data(p.1)[['treatment']])
```

# Add symptom onset date

Note that Patient 237 has the symptom onset date corrected in the .csv file (corrected to 9/27/21). For Patient 126, Acute date is corrected to 12/4/20 here.

```{r add metadata2}
# correct date for patient 126
sample_data(p.1)[['Date']][which(sample_data(p.1)[['ID']] == '126' & sample_data(p.1)[['Stage']] == 'A')] <- '12/4/20'



metatable <- as.data.frame(as.matrix(sample_data(p.1)))
meta2 <- read.csv('../data/R_input/V2_Demographics_August_2024_amended.csv') %>% 
  select(ID,Symptom.onset) %>%
  merge(.,select(metatable,c(sample.id.new,ID,Stage,Date)),by = 'ID', all.y = T) %>%
  mutate(
    Symptomonset.date = as.Date(Symptom.onset, format = "%m/%d/%y"),
    Measurement.date = as.Date(Date, format = "%m/%d/%y"),
    daysSinceSymptomOnset <- as.numeric(Measurement.date - Symptomonset.date)
  )


sample_data(p.1)[['Symptomonset.date']] = meta2$Symptomonset.date[match(sample_data(p.1)$sample.id.new,meta2$sample.id.new)]
sample_data(p.1)[['Measurement.date']] = meta2$Measurement.date[match(sample_data(p.1)$sample.id.new,meta2$sample.id.new)]
sample_data(p.1)[['daysSinceSymptomOnset']] = meta2$daysSinceSymptomOnset[match(sample_data(p.1)$sample.id.new,meta2$sample.id.new)]

```

# Add Omicron status

```{r add omicron}
sample_data(p.1)[['Omicron']] <- ifelse(is.na(sample_data(p.1)[['Symptomonset.date']]), NA,
                                         ifelse(sample_data(p.1)[['Symptomonset.date']]< as.Date("2022-01-01"), "Pre-Omicron", "Omicron"))
sample_data(p.1)[['Omicron']] = as.factor(sample_data(p.1)[['Omicron']])
```

# Final metadata

```{r remove redundant columns}
# remove redundant columns
redundant_cols = c('MiSeq_reads','Shannon_Dickson','Max_OTU_Dickson','rarefy_1000_Dickson','Date','read_count_raw')

for (i in redundant_cols){
  sample_data(p.1)[[i]] <- NULL
}

```

# Rarefy at 1,000 reads and perform TSS normalization (relative abundance)

```{r rarefy}
p.1r = rarefy_even_depth(p.1, sample.size = 1000, rngseed = T, replace = F, trimOTUs = TRUE, verbose = F)
p.1r.absolute <- p.1r
# TSS normalization (calculate relative abundance)
p.1r = transform_sample_counts(p.1r, function(x) 100*(x / sum(x))) 
metatable = as.data.frame(sample_data(p.1r))

p.1r.sampleCount = as.data.frame(table(metatable[['Stage']])) %>% transmute(group = Var1,SampleCount = Freq)
create_dt(p.1r.sampleCount,'Number of samples in each group after rarefaction at 1000 reads')
```

# Diversity calculation

```{r diversity calc}
alpha.div <- p.1r.absolute %>%
  estimate_richness %>%
    select(c(Observed,Shannon))
dist.wUF = UniFrac(p.1r, weighted=T, normalized=TRUE, parallel=F)

```

# Save RDS 

```{r save rds}
saveRDS(p.1,file.path(outdir,'phyloseqObj.specimen.deduplicated.updatedMetadata.unrarefied.rds'))
saveRDS(p.1r,file.path(outdir,'phyloseqObj.specimen.deduplicated.updatedMetadata.rarefied.rds'))
saveRDS(alpha.div,file.path(outdir,'alpha.diversity.rds'))
saveRDS(dist.wUF,file.path(outdir,'dist.weightedUniFrac.rds'))
```

# Session Information


```{r session info,eval=T}
sessionInfo()
save.image(file = file.path(outdir,"1.b.Metadata_rarefaction_diversityMetrics_calc.RData"))
