---
title: "Formal analysis -- 1.c Metadata summary"
date: "`r Sys.Date()`"
author: Yue (Gary) Zhang, PhD
output:
  rmdformats::readthedown:
    self_contained: true
    thumbnails: true
    lightbox: true
    gallery: false
    highlight: tango
    code_folding: hide
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = T, message = T,echo = T) 
```


```{r Initiate, results='hide'}
## initiate ---------------------------------------------------------
packages_to_load = c('ggplot2','vegan','ape','stringr','usedist','ggpubr','corrplot',
                     'cowplot','GGally','scales','metagMisc','ggsci','rstatix','wesanderson', 'readxl','tidyr',
                     'phyloseq','DT','plotly','ranacapa','ggrepel','purrr','dplyr','dendextend','lmerTest','Maaslin2',
                     'performance', 'ComplexHeatmap','circlize','openxlsx')
invisible(suppressPackageStartupMessages({
    lapply(packages_to_load, library, character.only = TRUE)
}))

#generating color palettes
col1 = wes_palette("Rushmore1")
col2 = wes_palette("GrandBudapest2")
col3 = wes_palette('Darjeeling2')
col21 <- rev(c("tomato1","darkblue","turquoise1","lightblue","darkred","mediumblue","purple","bisque",
               "greenyellow","yellow","violetred2","darkgreen","darkgoldenrod1","deeppink3","cadetblue4",
               "orchid2","seagreen3","purple4","dodgerblue2","red","gray27"))

set.seed(666)
```

```{r functions and image,eval=TRUE}
outdir = "../data/Results_publication_ready"
p.1r = readRDS(file.path(outdir,'phyloseqObj.specimen.deduplicated.correctedMetadata.rarefied.rds'))
p.all = readRDS(file.path(outdir,'phyloseqObj.all_samples.deduplicated.unrarefied.rds'))
allAnalysis.df = readRDS(file.path(outdir,'heatmap.metadata.rds'))

source('utils_functions.R')
source('auxiliaries.R')

```

# Read counts

Read count statistics of all specimens before rarefaction, including BAL (BAL-HC) specimens regardless of rarefaction status.

```{r read counts} 
p.1 = subset_samples(p.all,specimen_control == 'specimen')
p.1.readcount = as.data.frame(sample_sums(p.1)) 
names(p.1.readcount) <- 'num_reads'
stats_df <- data.frame(
  mean    = mean(p.1.readcount$num_reads),
  sd      = sd(p.1.readcount$num_reads),
  min     = min(p.1.readcount$num_reads),
  q1      = quantile(p.1.readcount$num_reads, 0.25),
  median  = median(p.1.readcount$num_reads),
  q3      = quantile(p.1.readcount$num_reads, 0.75),
  max     = max(p.1.readcount$num_reads),
  sum     = sum(p.1.readcount$num_reads)
)
rownames(stats_df) <- 'Number of reads'
create_dt(stats_df,'Read count statistics of all specimens before rarefaction')
```

# Rarefied

## Metadata summary

Results saved in excel spreadsheet named metadata.rarefied.summary.xlsx

```{r metadata, results='asis'}
metadata <- p.1r %>%
  sample_data %>%
  as.matrix %>%
  as.data.frame %>%
  mutate(Stage = factor(Stage,levels = Stages.HC.vec))

cat(paste0('Total number of patient (excluding HC) in after rarefaction: ',
           length(unique(metadata[metadata$Stage != 'Healthy Volunteers',]$ID))))


# generate another column named PEA avaibility for row annotation under this rule:
metadata$`Proteome` = ifelse(metadata$BAL_Protein_Cluster %in% c('Olink cluster 1',	'Olink cluster 2'), 'Available', 'Unavailable')


metadata.stage <- metadata %>%
  select(Stage, daysSinceSymptomOnset) %>%
  mutate(daysSinceSymptomOnset = as.numeric(daysSinceSymptomOnset)) %>%
  group_by(Stage) %>%
  summarise(
    sample_count = n(),
    daysSinceSymptomOnset_median = median(daysSinceSymptomOnset, na.rm = TRUE),
    daysSinceSymptomOnset_q25 = quantile(daysSinceSymptomOnset, 0.25, na.rm = TRUE),
    daysSinceSymptomOnset_q75 = quantile(daysSinceSymptomOnset, 0.75, na.rm = TRUE),
    daysSinceSymptomOnset_range_min = min(daysSinceSymptomOnset,na.rm = TRUE),
    daysSinceSymptomOnset_range_max = max(daysSinceSymptomOnset,na.rm = TRUE),
    .groups = "drop"
  )

metadata.stage.age.summary <- metadata %>%
  select(Stage, age) %>%
  mutate(age = as.numeric(age)) %>%
  group_by(Stage) %>%
  summarise(
    sample_count = n(),
    age_median = median(age, na.rm = TRUE),
    age_q25 = quantile(age, 0.25, na.rm = TRUE),
    age_q75 = quantile(age, 0.75, na.rm = TRUE),
    .groups = "drop"
  )

metadata.patient <- metadata %>%
  filter(specimen_type == 'BAL') %>%
  select(ID, age, gender, race,vaccination,Omicron,disease_severity, treatment) %>%
  mutate(age = as.numeric(age)) %>%
  unique

```
### Patient host-specific clinical variable summary

```{r acute sum}

table(metadata.patient$disease_severity)

table(metadata.patient$Omicron)

table(metadata.patient$vaccination)

table(metadata.patient$treatment)
```


```{r hc summary}
metadata.HC <- metadata %>%
  filter(specimen_type == 'BAL-HC') %>%
  select(ID, age, gender, race) %>%
  mutate(age = as.numeric(age)) 

cat('\n\n')
cat(sprintf(
  "Healthy volunteers age summary:\n  Median = %.1f\n  Mean = %.1f\n  SD = %.1f\n  IQR = %.1f–%.1f\n  Range = %d–%d\n",
  median(metadata.HC$age, na.rm = TRUE),
  mean(metadata.HC$age, na.rm = TRUE),
  sd(metadata.HC$age, na.rm = TRUE),
  quantile(metadata.HC$age, 0.25, na.rm = TRUE),
  quantile(metadata.HC$age, 0.75, na.rm = TRUE),
  min(metadata.HC$age, na.rm = TRUE),
  max(metadata.HC$age, na.rm = TRUE)
))
cat('\n\n')

cat('\n\n')
cat('HC gender')
cat('\n')
table(metadata.HC$gender)
```
```{r stage pairs}
patientPairs <- allAnalysis.df %>%
  filter(Stage != 'Healthy Volunteers' ) %>%
  mutate(Stage = factor(Stage,levels = Stages.vec)) %>%
  group_by(ID) %>%
  arrange(Stage) %>%
  summarise(
    Stage_new_combination      = paste(unique(Stage), collapse = " | "),
    daysSinceSymptomOnset_combination = paste(daysSinceSymptomOnset, collapse = " | "),
    Stage_new_pairs            = n_distinct(Stage),
    .groups = "drop"
  )

patientPairs.count <- patientPairs %>%
  group_by(Stage_new_combination) %>%
  summarise(Number_of_combination = n())

create_dt(patientPairs,'Stage-paris per patient')
create_dt(patientPairs.count,'Stage-paris counts')
```

```{r write metadata summary to xlsx}
# write.csv(metadata,file.path(outdir,'metadata_rarefied.csv'))
# metadata.stage %>% write.csv(file.path(outdir,'stage_nsample_median_quantiles_rarefied.csv'))
# metadata.stage.age.summary %>% write.csv(file.path(outdir,'stage_age_nsample_median_quantiles_rarefied.csv'))
# metadata.patient %>% write.csv(file.path(outdir,'patient_info_rarefied.csv'))
# metadata.HC %>% write.csv(file.path(outdir,'HealthyControl_info_rarefied.csv'))
# Create workbook
wb <- createWorkbook()

# 1) Metadata (full)
addWorksheet(wb, "metadata")
writeData(wb, "metadata", metadata)

# 2) Stage summary
addWorksheet(wb, "stage_summary")
writeData(wb, "stage_summary", metadata.stage)

# 3) Stage + age summary
addWorksheet(wb, "stage_age_summary")
writeData(wb, "stage_age_summary", metadata.stage.age.summary)

# 4) Patient info
addWorksheet(wb, "patient_info")
writeData(wb, "patient_info", metadata.patient)

# 5) HC info
addWorksheet(wb, "HealthyVolunteer_info")
writeData(wb, "HealthyVolunteer_info", metadata.HC)

# Save workbook
saveWorkbook(
  wb,
  file.path(outdir, "metadata.rarefied.summary.xlsx"),
  overwrite = TRUE
)

```


# Heatmap (Samples either with a ddPCR value or passed rarefaction)

Those are the BAL samples (i.e., excluding HC, as we don't show them in the dataset heatmap) we have included at least in one analysis (diversity and/or bacterial load).

```{r patient meta2, fig.width=12, fig.height=4}

# Step 1
m.p <- allAnalysis.df %>%
  filter(Stage != 'Healthy Volunteers') %>%
  distinct(ID, Stage,sample.id.new,rarefaction,ddpcr_available) %>%
  mutate(value = case_when(rarefaction == 'passed' & ddpcr_available == 'available' ~ 'ddPCR + 16S rRNA',
                           rarefaction == 'passed' & ddpcr_available == 'missing' ~ '16S rRNA-Only',
                           rarefaction == 'failed' & ddpcr_available == 'available' ~ 'ddPCR-Only',
                           TRUE ~ 'Not available')) %>%
  select(-c(sample.id.new,rarefaction,ddpcr_available)) %>% 
  unique()

# Make wide matrix: rows = Stage, columns = ID
mat <- m.p %>%
  pivot_wider(names_from = ID, values_from = value, values_fill = 'Not available') %>% # values_fill = Not available: fill missing combination with "Not available"
  tibble::column_to_rownames('Stage') %>%
  as.matrix()

# Convert matrix to long format to analyze per patient (column)
long_df <- as.data.frame(as.table(mat))
colnames(long_df) <- c("Stage", "ID", "value")

# For each ID, create a character vector of their stages
id_stage_combo <- long_df %>%
  filter(value != 'Not available') %>%
  mutate(Stage = factor(Stage,levels = Stages.vec)) %>%
  group_by(ID) %>%
  summarise(stage_combo = paste(sort(Stage), collapse = ",")) %>%
  ungroup()


# Define group order manually
group_priority <- c(
  "Acute,Recovery,Convalescent,Extended",
  "Acute,Recovery,Convalescent",
  "Recovery,Convalescent,Extended",
  "Acute,Recovery",
  "Recovery,Convalescent",
  "Acute,Convalescent",
  "Convalescent,Extended",
  "Acute",
  "Recovery",
  "Convalescent"
)

# Assign group number
id_stage_combo$group <- match(id_stage_combo$stage_combo, group_priority)

# Reorder IDs accordingly
ordered_IDs <- id_stage_combo %>%
  arrange(group) %>%
  pull(ID)

# Reorder columns of matrix
desired_stage_order <- Stages.vec

mat_ordered <- mat[desired_stage_order, ordered_IDs]

Heatmap(mat_ordered,
        name = "Sample",
        col = c("Not available" = "white","ddPCR-Only" = 'lightgreen', "16S rRNA-Only" = "orange", "ddPCR + 16S rRNA" = "lightblue"),
        cluster_rows = FALSE,
        cluster_columns = FALSE,
        column_title = "Patient",
        row_title = "",
        show_column_names = F,
        show_row_names = T
        )


# Clean and prepare annotation metadata
col_anno_df <- allAnalysis.df %>%
  filter(Stage != 'Healthy Volunteers') %>%
  distinct(ID, .keep_all = TRUE) %>%
  select(ID, vaccination, disease_severity, treatment, Omicron,BAL_Protein_Cluster) %>%
  mutate(
    Variant = Omicron,
    Proteome = ifelse(BAL_Protein_Cluster == 'Not included', 'Unavailable', 'Available')
  ) %>%
  filter(ID %in% colnames(mat_ordered)) %>%
  arrange(factor(ID, levels = colnames(mat_ordered)))

# sanity check for Proteome column
t.col <- col_anno_df %>% filter(Proteome == 'Available') %>% pull(ID)
t.meta <-metadata %>% filter(BAL_Protein_Cluster %in% c('Olink cluster 1', 'Olink cluster 2')) %>% pull(ID)
stopifnot(setequal(t.col,t.meta))

# sanity check for other columns selected (results: each ID only has one status for these selected columns.)
for (i in c('ID', 'vaccination', 'disease_severity', 'treatment', 'Omicron')) {
  stopifnot(length(unique(metadata$ID)) == length(unique(paste0(metadata[['ID']],metadata[[i]]))))
}
# check order
stopifnot(all(col_anno_df$ID == colnames(mat_ordered)))

top_anno <- HeatmapAnnotation(
  Vaccination = col_anno_df$vaccination,
  Severity    = col_anno_df$disease_severity,
  Treatment   = col_anno_df$treatment,
  Variant     = col_anno_df$Variant,
  `Proteome (acute)`    = col_anno_df$Proteome,
  col = list(
    Vaccination = colorscheme_vaccination,
    Severity    = colorscheme_severity,
    Treatment   = colorscheme_treatment,
    Variant     = colorscheme_omicron,
    `Proteome (acute)`    = colorscheme_proteome
  ),
  annotation_name_side = "right",
  show_annotation_name = TRUE,
  annotation_name_gp = gpar(fontsize = 10) 
)

(hm = Heatmap(mat_ordered,
        name = "Sample",
        col <- c("ddPCR + 16S rRNA"="#7BA4D9","16S rRNA-Only"="#EDB183","ddPCR-Only"="#8FBC8F","Not available"="#C8C8C8"),
        cluster_rows = FALSE,
        cluster_columns = FALSE,
        column_title = "Patient",
        row_title = "",
        show_column_names = FALSE,
        show_row_names = TRUE,
        row_names_gp = gpar(fontsize = 10),
        top_annotation = top_anno,
        border = "black",
       cell_fun = function(j, i, x, y, width, height, fill) {
         grid.rect(x = x, y = y, width = width, height = height,
                   gp = gpar(col = "black", fill = NA, lwd = 0.5))
         }))


saveplot(hm, 'heatmap.metadata.attributes.allSpecimens_ddpcr_or_rarefaction', w = 12, h = 3)
```