---
title: "Formal analysis -- 2.a Contamination check"
date: "`r Sys.Date()`"
author: Yue (Gary) Zhang, PhD
output:
  rmdformats::readthedown:
    self_contained: true
    thumbnails: true
    lightbox: true
    gallery: false
    highlight: tango
    code_folding: hide
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = T, message = T,echo = T) 
```


```{r Initiate, results='hide'}
## initiate ---------------------------------------------------------
packages_to_load = c('ggplot2','vegan','ape','stringr','usedist','ggpubr','corrplot',
                     'cowplot','GGally','scales','metagMisc','ggsci','rstatix','wesanderson', 'readxl','tidyr',
                     'phyloseq','DT','plotly','ranacapa','ggrepel','purrr','dplyr','dendextend','lmerTest','Maaslin2',
                     'performance', 'ComplexHeatmap','circlize','forcats','ggalluvial')
invisible(suppressPackageStartupMessages({
    lapply(packages_to_load, library, character.only = TRUE)
}))

#generating color palettes
col1 = wes_palette("Rushmore1")
col2 = wes_palette("GrandBudapest2")
col3 = wes_palette('Darjeeling2')
col21 <- rev(c("tomato1","darkblue","turquoise1","lightblue","darkred","mediumblue","purple","bisque",
               "greenyellow","yellow","violetred2","darkgreen","darkgoldenrod1","deeppink3","cadetblue4",
               "orchid2","seagreen3","purple4","dodgerblue2","red","gray27"))

set.seed(666)
```

```{r functions and image,eval=TRUE}
outdir = "../data/Results_publication_ready"
p.all = readRDS(file.path(outdir,'phyloseqObj.all_samples.deduplicated.unrarefied.rds'))

source('utils_functions.R')
source('auxiliaries.R')
```

# What are the read count comparing water blanks to samples

```{r blank read count ,fig.height=6, fig.width=8}

metatable = as.data.frame(as.matrix(sample_data(p.all)))
metatable$read_count_raw = as.numeric(metatable$read_count_raw)

metatable <- metatable %>%
  mutate(
    specimen_type = case_when(
      specimen_type %in% c("water","iso","AE","blank") ~ "Blanks",
      specimen_type == "mock"                         ~ "Mock community",
      specimen_type %in% c("BAL-HC","BAL")             ~ "BAL specimens",
      TRUE                                             ~ specimen_type
    ),
    specimen_type = factor(
      specimen_type,
      levels = c("BAL specimens", "Mock community", "Blanks"),
      labels = c("specimen", "mock community", "blank")
    )
  )


ggviolin(metatable, 
         x = "specimen_type", 
         y = "read_count_raw", 
         fill = "specimen_type",       
         palette = "jco",               
         add = "boxplot",              
         add.params = list(fill = "white", width = 0.2),
         trim = FALSE) +                # don't trim tails
  scale_y_log10(
    breaks = 10^seq(0, 7, by = 2),
    labels = trans_format("log10", math_format(10^.x))
  ) +                   
  labs(x = " ", y = "Read Count (log10)", 
       title = " ") +
  theme_pubr(base_size = 14) +    
  theme(axis.text.x = element_text(angle = 20, hjust = 1),
        legend.position = "none") +
  geom_hline(yintercept = 1000, linetype = "dashed", color = "darkred", linewidth = 0.8) 

 ggsave(file.path(outdir,'violinPlot.readCount.specimen_and_blanks.pdf'),width = 6,height = 6)

create_dt(summarize_grouped_stats(metatable,'specimen_type','read_count_raw'),
          'Read count summary')
```

## How many blanks do we have


```{r blanks sample count}
table(metatable$specimen_control)

```

## What are the top genera in blanks

Here we look for the top genera (in terms of average relative abundance) in blanks and in specimen from the unrarefied data.

The result suggests that Escherichia/Shigella, Staphylococcus, Corynebacterium, Anoxybacillus, Enterococcus, and Listeria are very likely to be contaminants.

```{r blank vs sample DA,fig.height=12, fig.width=15}
psq.tmp = p.all %>% subset_samples(., specimen_control %in% c('specimen','control'))

pheno <- psq.tmp %>%
  sample_data %>%
  as.matrix %>%
  as.data.frame %>% 
  select(c("sample.id.new","specimen_control", "read_count_raw"))

specimen.id <- pheno %>%
  filter(specimen_control == 'specimen') %>%
  pull(sample.id.new)
blank.id <- pheno %>%
  filter(specimen_control == 'control') %>%
  pull(sample.id.new)

exps <- taxReadCount_collapse_top10_from_phyloseq(psq.tmp,'Genus',sample_list = sample_names(psq.tmp))[[1]] %>%
      as.matrix %>%
      as.data.frame %>%
      setrowname 
exps[] <- lapply(exps, function(x) as.numeric(as.character(x)))
exps = decostand(exps,method = 'total',MARGIN = 2)
exps <- exps[!grepl("unclassified", rownames(exps)), ]

exps.specimen.genus.mean.top20 <- exps %>%
  select(all_of(specimen.id)) %>%
  transmute(genus_mean = rowMeans(across(everything()))) %>%
  arrange(desc(genus_mean)) %>%
  slice_max(order_by = genus_mean, n = 10) %>%
  tibble::rownames_to_column("genus") %>%
  mutate(group = 'specimen')
  
exps.blank.genus.mean.top20 <- exps %>%
  select(all_of(blank.id)) %>%
  transmute(genus_mean = rowMeans(across(everything()))) %>%
  tibble::rownames_to_column("genus") %>%
  filter(genus %in% exps.specimen.genus.mean.top20$genus) %>%
  mutate(group = 'blank')

exps.merged.genus.mean.top20 = bind_rows(exps.specimen.genus.mean.top20,exps.blank.genus.mean.top20)
exps.merged.genus.mean.top20$group = factor(exps.merged.genus.mean.top20$group, levels = c('specimen','blank'))

# set color palette to match above
jco <- pal_jco("default")(3)
pal <- c(specimen = jco[1], blank = jco[3])  # <- blank is the 3rd color

p = ggbarplot(
  exps.merged.genus.mean.top20,
  x = "genus",
  y = "genus_mean",
  facet.by = 'group', ncol = 1, 
  fill = 'group',
  palette = pal,
  color = "group",
  sort.val = "none",                  # <- don't let ggbarplot reorder
  order = exps.specimen.genus.mean.top20$genus,                # <- our explicit order
  x.text.angle = 45,       # Optional: tilt x-axis labels,
  title = ' ',
  xlab = '',
  ylab = 'Relative abundance',
  legend = "none" 
)
p <- ggpar(
  p,
  font.title       = c(22, "bold"),
  font.subtitle    = c(18),
  font.x           = c(18, "bold"),
  font.y           = c(18, "bold"),
  font.xtickslab   = c(16, "bold"),
  font.ytickslab   = c(16, "bold"),
  font.legend      = c(16),
  font.legend.title= c(16, "bold")
)

# Bigger facet strip labels too
p + theme(strip.text = element_text(size = 18, face = "bold"))
ggsave(file.path(outdir,'barplot.top10Genera.specimensAndBlanks.pdf'),width = 6,height = 12)
```

```{r scatters,fig.height=12, fig.width=15}

# scatter plots for relative abundance of top 20 genera in blanks vs read counts
exps.blank.genus.mean.top20 <- exps %>%
  select(all_of(blank.id)) %>%
  transmute(genus_mean = rowMeans(across(everything()))) %>%
  arrange(desc(genus_mean)) %>%
  slice_max(order_by = genus_mean, n = 20) %>%
  tibble::rownames_to_column("genus") 
exps.blankTop20 <- exps %>% 
  subset(., rownames(.) %in% exps.blank.genus.mean.top20$genus) %>%
  as.matrix %>%
  t %>%
  as.data.frame %>%
  merge(pheno,.,by = 0)

exps.blankTop20.long <- exps.blankTop20 %>%
  pivot_longer(cols = 5:24, names_to = "genus", values_to = "relative_abundance") %>%
  mutate(read_count_raw_log10 = log10(as.numeric(read_count_raw)),
         genus = factor(genus,levels = exps.blank.genus.mean.top20$genus)) 

ggscatter(x = 'read_count_raw_log10',
          y = 'relative_abundance',
          color = 'read_count_raw_log10',
          data = exps.blankTop20.long,
          facet.by = 'genus',
          title = '',
          xlab = 'log10(Read count)',
          ylab = 'Relative abundance') + 
  scale_color_gradient(low = "grey", high = "darkred") +
  labs(color = "log10(Read count)") +
  theme(axis.title = element_text(size = 16, face = "bold"))
ggsave(file.path(outdir,'scatterplot.abundanceVSreadcount.top20blankGenera.pdf'),width = 12,height = 12)

```

