---
title: "Formal analysis -- 3.b Diversity: Stage comparison"
subtitle: "All stages and HC"
date: "`r Sys.Date()`"
author: Yue (Gary) Zhang, PhD
output:
  rmdformats::readthedown:
    self_contained: true
    thumbnails: true
    lightbox: true
    gallery: false
    highlight: tango
    code_folding: hide
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = T, message = T,echo = T) 
```


```{r Initiate, results='hide'}
## initiate ---------------------------------------------------------
packages_to_load = c('ggplot2','vegan','ape','stringr','usedist','ggpubr','corrplot',
                     'cowplot','GGally','scales','metagMisc','ggsci','rstatix','wesanderson', 'readxl','tidyr',
                     'phyloseq','DT','plotly','ranacapa','ggrepel','purrr','dplyr','dendextend','lmerTest','Maaslin2',
                     'performance', 'ComplexHeatmap','circlize','forcats','ggalluvial')
invisible(suppressPackageStartupMessages({
    lapply(packages_to_load, library, character.only = TRUE)
}))

#generating color palettes
col1 = wes_palette("Rushmore1")
col2 = wes_palette("GrandBudapest2")
col3 = wes_palette('Darjeeling2')
col21 <- rev(c("tomato1","darkblue","turquoise1","lightblue","darkred","mediumblue","purple","bisque",
               "greenyellow","yellow","violetred2","darkgreen","darkgoldenrod1","deeppink3","cadetblue4",
               "orchid2","seagreen3","purple4","dodgerblue2","red","gray27"))

set.seed(666)
```

```{r functions and image,eval=TRUE}
outdir = "../data/Results_publication_ready"
p.1r = readRDS(file.path(outdir,'phyloseqObj.specimen.deduplicated.correctedMetadata.rarefied.rds'))

source('utils_functions.R')
source('auxiliaries.R')
```

```{r beta div,eval=T}
dist.wUF = readRDS(file.path(outdir,'dist.weightedUniFrac.RDS'))
alpha.div = readRDS(file.path(outdir,'alpha.diversity.rds'))
```

# Remove duplicated recovery samples

```{r remove duplicate recovery samples}
samples.duplicate.recovery = readRDS(file.path(outdir,'duplicate_recovery_samples_to_remove.RDS'))
p.1r = prune_samples(setdiff(sample_names(p.1r),samples.duplicate.recovery$sample.id.new),p.1r)

# sanity check
meta_df <- as.data.frame(as.matrix(sample_data(p.1r)))
dup.check <- meta_df %>%
  filter(Stage != "Healthy Volunteers") %>%
  count(ID, Stage, name = "n") %>% # Counts how many rows exist for each combination
  filter(n > 1)
```


# Differential abundance analysis

## Compare each disease stage against healthy volunteers

Here we set HV as the reference level, with including sex, age, and race as covariates. 

```{r stage VS hc diff}
psq.tmp = p.1r 
feature = 'Genus'
res_dir = file.path(outdir,'maaslin2','Stages_vs_HC')
regenerate_dir(res_dir)
input.tmp = prep_maaslin2_input(psq.tmp = psq.tmp, feature = feature, res_dir = res_dir)
exps = input.tmp[[1]]
pheno = input.tmp[[2]]

fit1 = capture.output(
  Maaslin2(input_data = exps, # capture.out is used to mutate verbose output by maaslin
           input_metadata = pheno,
           min_prevalence = 0.5, # the minimum percent of samples for which a feature is detected at minimum abundance (default 0.1)
           min_abundance = 0.05, # The minimum abundance for each feature (default 0)
           normalization = 'None', # default TSS
           transform = 'LOG', # default log 
           fixed_effects = c('Stage','race','gender','age'),
           reference = c('Stage,Healthy_Volunteers'),
           cores = 1,
           max_significance = 0.3, # this is set to high value so that maaslin automatically plot every figure that might need in future, #FDR q value threshold
           output = res_dir))
res <- read.csv(paste(res_dir, 'all_results.tsv', sep = '/'), sep = '\t') %>%
  filter(metadata == 'Stage') %>%
  group_by(value) %>%
  mutate(qval = p.adjust(pval, method = 'BH')) %>%
  ungroup()
write.table(res,paste(res_dir, 'all_results_filtered_by_Stage.tsv', sep = '/'), sep = '\t', quote = F, row.names = F)

create_dt(exps,'Maaslin2 input1: feature table')
create_dt(pheno,'Maaslin2 input2: phenotype table')
create_dt(res %>% filter(qval < 0.05),'Maaslin2 results (q <0.05)')

plt = plot_lfc(res,0.05,colorpalette = colorscheme)
print(plt)
ggsave(file.path(outdir,'forestPlot.stageVSHC.q0.05.pdf'),width = 5,height = 5)

```

# Fit Genus relative abundance to Weighted UniFrac PCoA (copied from 3.Diversity_analysis.Rmd)

```{r tax envfit}
grp = 'Stage'
dm = dist.wUF
metatable <- p.1r %>%
  sample_data %>%
  as.matrix %>%
  as.data.frame
metatable[['Stage']] = factor(metatable[['Stage']],levels = Stages.HC.vec)

dm_name = 'Weighted UniFrac'
envfitDf = tax_agg(p=p.1r)
topN_envfit_variables_to_plot = 3 # plot only arrows with FDR < 0.05, select arrows by R2. Select top3 which has r^2 > 0.3

env.taxa = plot_envifit(dm = dm,metatable = metatable, 
                        grp = grp ,dm_name = dm_name, 
                        envfitDf = envfitDf, topN_envfit_variables_to_plot = topN_envfit_variables_to_plot,
                        custom_colors = colorscheme,arrowshrinkingby = 0.5)

# print pcoa plot with gg_text_repel as a still image
print(env.taxa[[1]])
ggsave(filename = file.path(outdir,"PCoA.envFit.weightedUF.BAL+BALHC.pdf"), plot = env.taxa[[1]], width = 5, height = 5.5)
create_dt(env.taxa[[4]],'EnvFit')
```

# Permanova and Permdisp tests {.tabset}

## PCoA

```{r pcoa plot}
## weighted unifrac ---------------
grp = 'Stage'
dm = dist.wUF
metatable <- p.1r %>%
  sample_data %>%
  as.matrix %>%
  as.data.frame
metatable[['Stage']] = factor(metatable[['Stage']],levels = Stages.HC.vec)
dm_name = 'Weighted UniFrac'

cat('\n\n')
cat('### PCoA plot')
cat('\n\n')

plt = plot_unifrac(dm = dm,metatable = metatable, grp = grp,dm_name = dm_name,permtest = T,custom_colors = colorscheme)

```

## Compare stages pairwisely with blocking by patients

```{r}
# test results 
## paired tests ---
cat('\n\n')
cat('### Permutational test raw output (compare stages; paired by patientID)')
cat('\n\n')
perm.res <- lapply(stageTest.paired, function(groups) {
  perm_test.blocked(metatable, dm, groups)
})
cat('\n\n')
cat('### Permutational test p value summary (compare stages; paired)')
cat('\n\n')
perm.res.df <- do.call(rbind, lapply(perm.res, function(x) {
  data.frame(comparison = x[1], permanova_p_value = x[2], permdisp_p_value = x[3], patient.count = x[4], stringsAsFactors = FALSE)
}))
create_dt(perm.res.df,'Summary of permutational analysis p values')
write.csv(perm.res.df,file.path(outdir,'weightedUnifrac.permTests.stagePairs.blockByPatients.csv'))
```

## Compare every disease stage to HC {.tabset}

```{r perm 2}
## unpaired tests ---
cat('\n\n')
cat('### Permutational test raw output (compare HC to each stage; unpaired)')
cat('\n\n')
perm.res <- lapply(stageTest.unpaired, function(groups) {
  perm_test(metatable, dm, groups)
})

cat('\n\n')
cat('### Permutational test p value summary (compare HC to each stage; unpaired)')
cat('\n\n')
perm.res.df <- do.call(rbind, lapply(perm.res, function(x) {
  data.frame(comparison = x[1], permanova_p_value = x[2], permdisp_p_value = x[3], stringsAsFactors = FALSE)
}))

create_dt(perm.res.df,'Summary of permutational analysis p values')
write.csv(perm.res.df,file.path(outdir,'weightedUnifrac.permTests.stagesVSHC.csv'))
```


# Top genera relative abundance alluvium plot

## Top genera across all samples

note that Ecoli is removed. unclassified genera are merged into others, which is not shown.

```{r alluvian style2,fig.height=20, fig.width=18}
psq.tmp = p.1r

pheno <- psq.tmp %>%
  sample_data %>%
  as.matrix %>%
  as.data.frame %>% 
  select(c("sample.id.new","Stage"))

id.a <- pheno %>%
  filter(Stage == 'Acute') %>%
  pull(sample.id.new)
id.r <- pheno %>%
  filter(Stage == 'Recovery') %>%
  pull(sample.id.new)
id.c <- pheno %>%
  filter(Stage == 'Convalescent') %>%
  pull(sample.id.new)
id.e <- pheno %>%
  filter(Stage == 'Extended') %>%
  pull(sample.id.new)
id.h <- pheno %>%
  filter(Stage == 'Healthy Volunteers') %>%
  pull(sample.id.new)

exps <- taxReadCount_collapse_topn_from_phyloseq(psq.tmp,'Genus',sample_list = sample_names(psq.tmp),topn = 11,merge_unclassified = T)[[2]] %>%
      as.matrix %>%
      as.data.frame %>%
      setrowname 
exps[] <- lapply(exps, function(x) as.numeric(as.character(x)))
exps = exps/100 # originally each column adds up to 100
exps = exps[-nrow(exps),] # remove the "Others" row
exps = exps[-which(rownames(exps) == "Escherichia/Shigella"),]

# 1) Means
id.a.mean <- rowMeans(exps[, id.a, drop = FALSE], na.rm = TRUE)
id.r.mean <- rowMeans(exps[, id.r, drop = FALSE], na.rm = TRUE)
id.c.mean <- rowMeans(exps[, id.c, drop = FALSE], na.rm = TRUE)
id.e.mean <- rowMeans(exps[, id.e, drop = FALSE], na.rm = TRUE)
id.h.mean <- rowMeans(exps[, id.h, drop = FALSE], na.rm = TRUE)

mean_df <- tibble(
  genus = rownames(exps),
  Acute = id.a.mean,
  Recovery = id.r.mean,
  Convalescent = id.c.mean,
  Extended = id.e.mean,
  `Healthy Volunteers` = id.h.mean,
)

# 2) Group-specific orders (within those 10 genera)
order.a <- mean_df %>%
  arrange(desc(Acute)) %>%
  pull(genus)
order.r <- mean_df %>%
  arrange(desc(Recovery)) %>%
  pull(genus)
order.c <- mean_df %>%
  arrange(desc(Convalescent)) %>%
  pull(genus)
order.e <- mean_df %>%
  arrange(desc(Extended)) %>%
  pull(genus)
order.h <- mean_df %>%
  arrange(desc(`Healthy Volunteers`)) %>%
  pull(genus)

# 4) Long data + group-specific stratum factor
plot_df <- mean_df %>%
  pivot_longer(Stages.HC.vec, names_to = "group", values_to = "abundance") %>%
  mutate(
    group = factor(group, levels = Stages.HC.vec),
    # make a group-specific stratum label so each side can have its own order
    stratum_id = paste0(group,genus)
  )

# set stratum levels: 
plot_df$stratum_id <- factor(plot_df$stratum_id, 
                             levels = c(paste0('Acute',order.a),
                                        paste0('Recovery',order.r),
                                        paste0('Convalescent',order.c),
                                        paste0('Extended',order.e),
                                        paste0('Healthy Volunteers',order.h)))

# 5) Plot 
# Show a label for each block ( number only (x%))
plot_df <- plot_df %>%
  mutate(genus = factor(genus, levels = order.a),
         genus_perc = paste0(round(abundance * 100, 0), "%"))
create_dt(plot_df,'Top 10 genera in specimen')
ggplot(plot_df,
       aes(x = group,
           stratum = stratum_id,      # <- group-specific ordering
           alluvium = genus,          # <- links the same genus across sides
           y = abundance,
           fill = genus)) +
  scale_x_discrete(expand = c(.1, .1)) +
  geom_flow(alpha = 0.6,width = 0.6) +
  geom_stratum(alpha = 0.85, color = "grey20", width = 0.6) +
  scale_y_continuous(labels = percent_format(accuracy = 0.1)) +
  theme_minimal(base_size = 12) +
  scale_fill_manual(values = col21) +   
  xlab('') +
  ylab('Relative abundance') +
  theme(panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      panel.background = element_blank(),
      axis.line = element_line(colour = 'black'),
      legend.position = "bottom",          # place legend on the right
      legend.title = element_blank(),     # optional: no title
      legend.text = element_text(size = 22,face = "bold"),
      axis.text = element_text(size = 22,face = 'bold', colour = "black"),
      axis.text.x = element_text(angle = 40, size = 22, vjust = 1, hjust = 1, face = "bold",colour = "black"),
      axis.title = element_text(size = 22, face = "bold", colour = "black")) +
  guides(fill = guide_legend(nrow = 5, byrow = TRUE)) +
  ggtitle("Top 10 Genera selected by average across all samples",
          subtitle = "Ecoli is removed. unclassified genera are merged into others, which is not shown.") +
  geom_text(
    stat = "stratum",
    aes(label =  ifelse(abundance > 0.005, as.character(genus_perc), '')),   # <-- cleaned name
    size = 10,
    color = "black"
  )
ggsave(file.path(outdir,'Alluvium.top10GeneraInAllSamples.StagesAndHC.percOnly.pdf'),width = 12,height = 18)


```


# Alpha diversity

```{r alpha}
metatable <- p.1r %>%
  sample_data %>%
  as.matrix %>%
  as.data.frame

alpha.div.long <- metatable %>%
  select(ID,Stage) %>%
  merge(.,alpha.div,by=0,all.x=T) %>%
  pivot_longer(cols = c('Observed','Shannon'), # Columns to stack
               names_to = "Metric",                           # New column for metric names
               values_to = "Value") %>%
  dplyr::mutate(Stage = factor(Stage,levels = Stages.HC.vec)) %>%
  dplyr::mutate(Metric = ifelse(Metric == 'Observed', 'Observed OTUs',
                                ifelse(Metric == 'Shannon', 'Shannon Index','ERROR')))

alpha.div.long %>%
  ggplot(aes(x = Stage, y = Value, fill = Stage)) +
  geom_violin(trim = FALSE) +
  geom_boxplot(width = 0.2, fill = "white", outlier.size = 0.5) +
  facet_wrap(~Metric, scales = "free_y") +
  scale_fill_manual(values = colorscheme) +
  stat_compare_means(
    method = "wilcox.test",
    comparisons = stageTest.unpaired,   
    label = "p.signif",            
    color = "darkred",
    size = 6
  ) +
  labs(x = " ", y = "Metric value", title = " ") +
  theme_pubr(base_size = 14) +
  theme(
    axis.text.x = element_text(angle = 20, hjust = 1,face = 'bold',size = 15),
    axis.text.y = element_text(angle = 20, hjust = 1,face = 'bold',size = 15),
    axis.title.y = element_text(face = 'bold',size = 20),
    legend.position = "none"
  )
ggsave(file.path(outdir,'violinPlot.alpha.Stages_vs_HC.pdf'),width = 10,height = 8)

alpha.stat <- compare_means(
  Value ~ Stage, 
  data = alpha.div.long, 
  group.by = 'Metric',
  method = "wilcox.test", 
  comparisons = stageTest.unpaired) %>% 
  filter(group2 == 'Healthy Volunteers') 

alpha.stat %>%
  create_dt(.,'Alpha diversity statistical comparisons')
alpha.stat %>%
  write.csv(.,file.path(outdir,'alphaDiversity.stats.stagesVSHC.csv'))
```

# Session information

```{r sess info}
sessionInfo()

```