---
title: "multi-modal data integration"
subtitle: "sPLS-based analysis"
date: "`r Sys.Date()`"
author: Yue (Gary) Zhang, PhD
output:
  rmdformats::readthedown:
    self_contained: true
    thumbnails: true
    lightbox: true
    gallery: false
    highlight: tango
    code_folding: hide
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = T, message = T,echo = T) 
```


```{r Initiate, results='hide'}
.libPaths('/Users/zhangy68/R/R_4.4.2')
## initiate ---------------------------------------------------------
packages_to_load = c('ggplot2','vegan','ape','stringr','usedist','ggpubr','corrplot',
                     'cowplot','GGally','scales','metagMisc','ggsci','rstatix','wesanderson', 'readxl','tidyr',
                     'phyloseq','DT','plotly','ranacapa','ggrepel','purrr','dplyr','dendextend','lmerTest','Maaslin2',
                     'performance','mixOmics','ComplexHeatmap','circlize')
invisible(suppressPackageStartupMessages({
    lapply(packages_to_load, library, character.only = TRUE)
}))

#generating color palettes
col1 = wes_palette("Rushmore1")
col2 = wes_palette("GrandBudapest2")
col3 = wes_palette('Darjeeling2')
col21 <- rev(c("tomato1","darkblue","turquoise1","lightblue","darkred","mediumblue","purple","bisque",
               "greenyellow","yellow","violetred2","darkgreen","darkgoldenrod1","deeppink3","cadetblue4",
               "orchid2","seagreen3","purple4","dodgerblue2","red","gray27"))

set.seed(666)
```

```{r functions and image,eval=TRUE}
indir = "../../data/R_input"
outdir = "../../data/Formal_analysis_files2"
load(file.path(outdir,'3.c.Diversity.clinical_effects.rdata'))
# overwrite dir path after loading rdata
indir = "../../data/R_input" # overwrite dir path after loading rdata
outdir = "../../data/Formal_analysis_files2" # overwrite dir path after loading rdata

source('../utils_functions.R')
set.seed(666)
```

```{r mixomics specific functions}

# remove low abundance OTU for mixOmics analysis
low.count.removal <- function(
    data, # OTU count data frame of size n (sample) x p (OTU)
    percent=0.01 # cutoff chosen
) 
{
  keep.otu = which(colSums(data)*100/(sum(colSums(data))) > percent)
  data.filter = data[,keep.otu]
  return(list(data.filter = data.filter, keep.otu = keep.otu))
}

mixomics.CLR = function(dataset,removeLowOtu = T, lowOtuCutOff=0.01){
  if(removeLowOtu){
    # call the function then apply on the offset data
    result.filter <- low.count.removal(dataset, percent=lowOtuCutOff)
    dataset <- as.matrix(result.filter$data.filter)
  }
  # undergo CLR transformation
  ## apply offset
  dataset <- dataset + 0.01
  data.clr = decostand(dataset,MARGIN = 1, method = 'clr') %>% as.data.frame
  return(data.clr)
}

mixOmics.PCA = function(df,group,ncomp=10,scale=F,title){
  # # test variables
  # df <- data.clr
  # group <- predictor
  # ncomp = 10
  # scale = F
  # title = 'test'
  
  # tuning PCA model in terms of how many components to include
  tune.pca.multi <- tune.pca(df, ncomp = 10, scale = scale)
  plot.tune = plot(tune.pca.multi)
  
  final.pca.multi <- pca(df, ncomp = ncomp,scale = scale)
  
  plot.pca = plotIndiv(final.pca.multi,
            comp = c(1, 2),   # Specify components to plot
            ind.names = TRUE, # Show row names of samples
            group = group,
            title = title,
            legend = TRUE, legend.title = '')
  plot.var = plotVar(final.pca.multi, comp = c(1, 2),
                     var.names = TRUE,
                     cex = 3,         # To change the font size
                     # cutoff = 0.5,  # For further cutoff
                     title = title)
  plot.bi = biplot(final.pca.multi, group =group, 
                   legend.title = ' ')
  return(list(plot.tune,plot.pca,plot.var,plot.bi))
}

mixOmics.PLSDA = function(df,group,ncomp=10,title,scale = F){
  # # test variables
  # df <- data.clr
  # group = predictor
  # ncomp = 10
  # title = 'test'
  # scale = F
  
  plsda <- plsda(df,group, ncomp = 10,scale = scale)
  perf.plsda <- perf(plsda, validation = 'Mfold', folds = 3, 
                     progressBar = FALSE,  # Set to TRUE to track progress
                     nrepeat = 10)         # We suggest nrepeat = 50
  
  plot.error = plot(perf.plsda, sd = TRUE, legend.position = 'horizontal')
  plsda <- plsda(df,group, ncomp = ncomp,scale = scale)
  plot.plsda.1_2 = plotIndiv(plsda, ind.names = FALSE, legend=TRUE,
                             comp=c(1,2), ellipse = TRUE, 
                             title = paste0(title,': comp 1-2'),
                             X.label = 'PLS-DA comp 1', Y.label = 'PLS-DA comp 2')
  
  plot.plsda.2_3 = plotIndiv(plsda, ind.names = FALSE, legend=TRUE,
                             comp=c(2,3), ellipse = TRUE, 
                             title =  paste0(title,': comp 2-3'),
                             X.label = 'PLS-DA comp 2', Y.label = 'PLS-DA comp 3')
  
  return(list(plot.error,plot.plsda.1_2,plot.plsda.2_3,plsda))
}

plot_unifrac_longXaxis = function(dm,metatable,grp,dm_name,permtest = F,grp_continuous = F,custom_colors){
  # dummy input variable for testing function
  # dm = distmtx
  # metatable = metadata.o
  # dm_name = 'x'
  # grp = grouping
  
  
  metatable = subset(metatable, get(grp) != 'N.A.', select = c(grp))
  # select distance matrix matching to metadata
  dm = dist_subset(dm,rownames(metatable))
  # some distance measures may result in negative eigenvalues. In that case, add a correction:
  pcoa.wunifrac = pcoa(dm,correction = 'cailliez')
  # transform the pcoa.wunifrac object into a df for making it plottable with ggplot2
  coordinates.pcoa.wunifrac = data.frame(pcoa.wunifrac$vectors,check.names = FALSE)
  
  # Adonis test (only run if all groups have >= 5 samples)
  ## Calculate the number of samples in each group
  group_counts <- table(metatable[[grp]])
  ## Check if any group has fewer than 5 samples
  any_small_groups <- any(group_counts < 5)
  if (any_small_groups == F & permtest ==T){
    stopifnot(all(rownames(metatable) == names(dm)))
    adonis.r = adonis2(dm~get(grp),metatable,permut = 1000,parallel = 1)
    adonis.p = round(adonis.r$`Pr(>F)`[1],digit = 4)
    
    betadisp.r = permutest(betadisper(dm,metatable[[grp]]),permutations = 1000,parallel = 1)
    betadisp.p = round(betadisp.r$tab[[6]][1],digit = 4)
    
    subtitle = paste0('permanova p val = ', adonis.p, ',betadisp p val = ', betadisp.p)
  } else {
    subtitle = 'Adonis/permdisp skipped'
  }
  
  
  # plot
  if (!all(rownames(metatable) == rownames(coordinates.pcoa.wunifrac))) {stop("mismatchs in pca coordinates and coloring column")} # sanity check
  
  # Calculate group centroids
  coordinates = coordinates.pcoa.wunifrac[,c(1,2)]
  #coordinates[[grp]]= metatable[[grp]]      
  coordinates[[grp]] = metatable[[grp]][match(rownames(coordinates),rownames(metatable))] 
  #print(coordinates)
  centroids <- coordinates %>%
    group_by_at(grp) %>%
    summarize(CentroidX = mean(Axis.1), CentroidY = mean(Axis.2))
  coordinates = merge(coordinates,centroids,
                      by = grp)           
  # automatically grab relative eigenvalue
  if ("Rel_corr_eig" %in% colnames(pcoa.wunifrac$values)) {
    pc_exp <- pcoa.wunifrac$values$Rel_corr_eig
  } else {
    pc_exp <- pcoa.wunifrac$values$Relative_eig
  }
  
  if (grp_continuous == F){
    
  pcoa <- ggplot() +
  # Smaller scatter points
  geom_point(data = coordinates,
             mapping = aes(x = `Axis.1`, y = `Axis.2`, colour = get(grp)),
             alpha = 0.6, size = 0.5) + 
  
  # Connection lines to centroids
  geom_segment(data = coordinates,
               aes(x = CentroidX, xend = Axis.1, y = CentroidY, yend = Axis.2, color = get(grp)),
               alpha = 0.3, size = 0.3) +
  # Add ellipses
  #stat_ellipse(data = coordinates,
  #              aes(x = `Axis.1`, y = `Axis.2`, color = get(grp)),
  #               type = "norm", level = 0.68, size = 0.5, alpha = 0.4) +
  # Emphasize centroids
  geom_point(data = centroids,
             aes(x = CentroidX, y = CentroidY, fill = get(grp)),
             size = 4, shape = 21, stroke = 1.2, color = 'black') +
  
  labs(title = paste0(dm_name, ': ', grp),
       subtitle = subtitle) +
  
  xlab(paste('PC1 (', label_percent()(pc_exp[1]), ')', sep = '')) +
  ylab(paste('PC2 (', label_percent()(pc_exp[2]), ')', sep = '')) +
  xlim(-0.5,0.3) +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        axis.line = element_line(colour = 'black'),
        legend.position = "top",          # place legend on the right
        legend.title = element_blank(),     # optional: no title
        legend.text = element_text(size = 8,face = "bold"),
        axis.text = element_text(size = 12),
        axis.title = element_text(size = 14, face = "bold")) +
  # Apply custom color scheme
  scale_color_manual(values = custom_colors)+
  scale_fill_manual(values = custom_colors)
  
  } else {
    pcoa <- ggplot() +
      # Smaller scatter points
      geom_point(data = coordinates,
                 mapping = aes(x = `Axis.1`, y = `Axis.2`, colour = !!sym(grp)),
                 alpha = 0.6, size = 0.5) + 
      
      labs(title = paste0(dm_name, ': ', grp),
           subtitle = subtitle) +
      
      xlab(paste('PC1 (', label_percent()(pc_exp[1]), ')', sep = '')) +
      ylab(paste('PC2 (', label_percent()(pc_exp[2]), ')', sep = '')) +
      
      theme(panel.grid.major = element_blank(),
            panel.grid.minor = element_blank(),
            panel.background = element_blank(),
            axis.line = element_line(colour = 'black'),
            legend.position = "top",          # place legend on the right
            legend.title = element_blank(),     # optional: no title
            legend.text = element_text(size = 8,face = "bold"),
            axis.text = element_text(size = 12),
            axis.title = element_text(size = 14, face = "bold")) +
      # Apply custom color scheme
      scale_color_manual(values = custom_colors)+
      scale_fill_manual(values = custom_colors)

    
      
  #   pcoa <- ggplot() +
  #     geom_point(data = coordinates,
  #                mapping = aes(x = `Axis.1`, y = `Axis.2`, colour = !!sym(grp)),
  #                alpha = 0.8, size = 1) + 
  #     labs(title = paste0(dm_name, ': ', grp),
  #          subtitle = subtitle) +
  #     geom_segment(data = coordinates,
  #                  aes(x = CentroidX, xend = Axis.1, y = CentroidY, yend = Axis.2, color = !!sym(grp)),
  #                  alpha = 0.2, size = 0.5) +
  #     xlab(paste('PC1 (', label_percent()(pc_exp[1]), ')', sep = '')) +
  #     ylab(paste('PC2 (', label_percent()(pc_exp[2]), ')', sep = '')) +
  #     theme(panel.grid.major = element_blank(),
  #           panel.grid.minor = element_blank(),
  #           panel.background = element_blank(),
  #           axis.line = element_line(colour = 'black'),
  #           legend.title = element_blank(),
  #           legend.position = 'bottom',
  #           legend.text = element_text(size = 8)) +
  #     scale_color_gradient(low = "blue", high = "red")  # Adjust colors for your preference
  #   
  }
  
  print(pcoa)
}

# Save mixomics plots
saveplot = function(plt,filename,w = 5, h = 5) {
  pdf(file.path(outdir, paste0(filename,".pdf")), width = w, height = h)
  try({print(plt)}, silent = TRUE)
  dev.off()
}
```

# Microbiome associated with the olink clusters

## PCoA

```{r olink cluster, fig.height=8, fig.width=10}
colorscheme_BAL_Protein_Cluster = c(
  "Olink cluster 1" = "orange",
  "Olink cluster 2" = "darkblue",
  "Not included" = "grey"
)

p.tmp <- subset_samples(p.1r, Stage == 'Acute' & BAL_Protein_Cluster!= "Not included")
grp = 'BAL_Protein_Cluster'
dm = dist.wUF
metatable <- p.tmp %>%
  sample_data %>%
  as.matrix %>%
  as.data.frame

create_dt(metatable, 'Metadata of the samples for integration')

dm_name = 'BAL_Protein_Cluster in Acute samples'
plt = plot_unifrac_longXaxis(dm = dm,metatable = metatable, grp = grp,dm_name = dm_name,permtest = T,custom_colors = colorscheme_BAL_Protein_Cluster)
ggsave(file.path(outdir,'PCoA.OlinkClusters.pdf'),width = 4,height = 4)

alpha.div.long.extended %>%
  subset(Stage == 'Acute' & BAL_Protein_Cluster!= "Not included") %>%
  ggplot(aes(x = get(grp), y = Value, fill = get(grp))) +
  geom_violin(trim = FALSE) +
  geom_boxplot(width = 0.2, fill = "white", outlier.size = 0.5) +
  facet_wrap(~Metric, scales = "free_y") +
  scale_fill_manual(values = colorscheme_BAL_Protein_Cluster) +
  stat_compare_means(method = "wilcox.test", label = "p.signif", label.x.npc = "middle", label.y.npc = 'top',color='darkred',size = 10) +
  labs(x = " ", y = "Value", title = " ") +
  theme_pubr(base_size = 14) +
  theme(axis.text.x = element_text(angle = 20, hjust = 1),
        legend.position = "none")


 ggsave(file.path(outdir,'violinPlot.olinkClusters.alpha_groupwise.pdf'),width = 6,height = 8)


```

## Hierarchical clustering with the weighted UniFrac distance

Using Ward D clustering algorithm.

Ward D hierarchical clustering (also called Ward's method or Ward.D) is a type of agglomerative hierarchical clustering that minimizes the total within-cluster variance at each step. It is especially useful when you want compact clusters.

```{r hier,fig.width=10,fig.height=8, eval=FALSE}
dm = dist_subset(dm,rownames(metatable))
i = 'ward.D'
hc <- hclust(dm, method = i)

# Plot the dendrogram
dend <- as.dendrogram(hc)
tip.cluster = metatable$BAL_Protein_Cluster[match(labels(dend),rownames(metatable))]
colors <- ifelse(tip.cluster == "Olink cluster 1", "orange",'darkblue')

dend <- dend %>%
  set("labels_col", colors) %>%  # Apply colors
  set("labels_cex", 1)         # Optional: Adjust label size for visibility

# Plot the customized dendrogram
plot(dend, main = paste0("Dendrogram with ",i, ' approach'))

source('plot_hierClust.R')
```

# Preprocess integration data

## Microbiome part

Here we extract the rarefied OTU count matrix, remove low-abundance OTUs, and perform CLR transformation

Remove low-abundance OTUs: the sum of each OTU across all samples (24 samples in total) * 100 / the sum of all OTU across all samples (24 * 100) <= 0.01

CLR transformation (0.01 pseudo count added beforehand)

```{r preprocess}
phylo =  p.tmp
otu.data <- phylo %>%
  otu_table %>%
  as.matrix %>%
  t %>%
  as.data.frame
predictor = sample_data(phylo)[['BAL_Protein_Cluster']]
names(predictor) = rownames(sample_data(phylo))
table(predictor)
stopifnot(rownames(otu.data) == names(predictor))
data.clr = mixomics.CLR(otu.data)

# check the number of variables kept after filtering
cat('\n\n')
cat(paste0('Num. OTUs before filtering for low abundance OTUs: ', ncol(otu.data)))
cat('\n\n')
cat(paste0('Num. OTUs after filtering for removing low abundance OTUs: ',  ncol(data.clr)))
cat('\n\n')

```

## Olink data

```{r olink preprocess}
ol2 = read.csv(file.path(indir,'NPX_List_of_Acute_proteins_in_Olink.csv'), check.names = F)
ol2 = ol2[-1,-c(2:5)] # remove the subheader row and control column

# add suffix to rows with duplicated assay names (rowname)
ol2 <- ol2 %>%
  group_by(NIHID) %>%
  mutate(NIHID = if (n() == 1) NIHID else paste0(NIHID, "_", row_number())) %>%
  ungroup() %>%
  as.data.frame %>%
  setrowname

# change patient ID to sample.id.new to match to the otu table
names(ol2) = sample_data(phylo)$sample.id.new[match(names(ol2),sample_data(phylo)$ID)]

# convert all columns to numeric
ol2[] = apply(ol2,2,as.numeric) 
saveRDS(ol2,file.path(outdir,'ol2.rds'))

# reorder samples to match the otu table and transpose it
proteome.matrix <- ol2 %>%
  dplyr::select(rownames(as.data.frame(data.clr))) %>%
  as.matrix %>%
  t
# sanity check
stopifnot(all(rownames(proteome.matrix) == rownames(as.data.frame(data.clr))))

# remove the lowly expressed genes (skip this step)
proteome.matrix.filtered = proteome.matrix

# check the number of variables kept after filtering
cat('\n\n')
cat(paste0('Num. Features before filtering for low abundance: ', ncol(proteome.matrix)))
cat('\n\n')
cat(paste0('Num. Features after filtering for removing low abundance features: ',  ncol(proteome.matrix.filtered)))
cat('\n\n')

```

```{r pre-integration sanity check}
stopifnot(all(rownames(proteome.matrix.filtered) == rownames(as.data.frame(data.clr))))
stopifnot(all(rownames(proteome.matrix.filtered) == names(predictor)))
```


# sPLS-DA on Olink Data

Here we first attempt to find the most discriminant features that can separate samples into the two clusters using sPLS-DA

## Here we first verify the presence of the two clusters with PCA and PLS-DA

```{r pca plsda on olink}
olk.pca <- pca(proteome.matrix.filtered, ncomp = 2, scale = TRUE)
f1 = plotIndiv(olk.pca, group = predictor, ind.names = FALSE,
          legend = TRUE, 
          title = 'PEA PCA',
          style = 'ggplot2', star = TRUE,
          col = colorscheme_BAL_Protein_Cluster[1:2])

## PLS-DA
olk.plsda <- plsda(proteome.matrix.filtered,predictor, ncomp = 5) 
f2 = plotIndiv(olk.plsda, group = predictor, ind.names = FALSE,
          legend = TRUE, 
          title = 'PEA, PLSDA comp 1 - 2',
          style = 'ggplot2', star = TRUE,,
          col = colorscheme_BAL_Protein_Cluster[1:2])


perf.olk.plsda <- perf(olk.plsda, validation = 'Mfold', 
                               folds = 3, 
                               progressBar = FALSE, # TRUE to track progress
                               nrepeat = 50) # we recommend 50 
plot(perf.olk.plsda, sd = TRUE)

background.max <- background.predict(olk.plsda, 
                                     comp.predicted = 2,
                                     dist = 'max.dist') 

f3 = plotIndiv(olk.plsda, comp = 1:2, group = predictor,
          ind.names = FALSE, title = 'PEA PLSDA',
          legend = TRUE,  background = background.max,
          style = 'ggplot2', star = TRUE,,
          col = colorscheme_BAL_Protein_Cluster[1:2])
saveplot(f1,'PCA.olink',w=6)
saveplot(f3,'PLSDA.olink',w=6)
```

The axis labels indicate the amount of variation explained per component. However, the interpretation of this amount is not as important as in PCA, as PLS-DA aims to maximise the covariance between components associated to X and Y, rather than the variance X.

The results show a distinct clustering pattern.


## Heatmap of all proteins

```{r heatmap all proteins,fig.height=15,fig.width=10,eva=F}
# PEA heatmap
## Main heatmap
mat = t(olk.plsda$X)
cat(paste0('NPX matrix min and max are: ', round(min(mat),digits = 2), ' and ', round(max(mat),digits = 2)))
## Add loading weightes on the right side as an one-column heatmap
loading_weight_vector <- olk.plsda$loadings$X %>%
  as.data.frame %>%
  dplyr::select(comp1) %>%
  as.matrix()
loading_weight_vector <- loading_weight_vector[rownames(mat), , drop = FALSE]
min(loading_weight_vector)
max(loading_weight_vector)
col_fun_loading <- colorRamp2(c(-0.05, 0, 0.03), c("blue", "white", "red"))

htmp.loading.annot.PEA <- rowAnnotation(`Loading weights` = loading_weight_vector,col = list(`Loading weights` = col_fun_loading),show_annotation_name = FALSE)


col_fun = colorRamp2(c(-5, 0, 5), c("#2B5C8AFF", "#FFF6EDFF", "#AC4522FF"))
hm = Heatmap(mat, name = 'PEA full set',
             column_title = "",
             show_row_names = F,show_row_dend = T,
             row_dend_width = unit(30,'mm'),
             column_dend_height = unit(20,'mm'),
             show_column_names = T, show_column_dend = T,
             right_annotation = htmp.loading.annot.PEA,
             cluster_columns = T,
             col = col_fun,
             border_gp = gpar(col = "black", lty = 1),
             clustering_method_rows = 'ward.D',
             row_names_max_width = unit(10,'cm')
)

# Add the outcome variable (olink cluster) as an one-row heatmap at the bottom
pred.vector <- matrix(predictor,
                      nrow = 1, 
                      dimnames = list("PEA Cluster", names(predictor)))
stopifnot(all(colnames(mat) == colnames(pred.vector)))

htmp.colAnnot <- Heatmap(pred.vector,
                         name = 'PEA Cluster',
                         show_row_names = T,
                         show_row_dend = FALSE,
                         show_column_names = FALSE,
                         show_column_dend = FALSE,
                         col = colorscheme_BAL_Protein_Cluster,
                         height = unit(8, "mm"),
                         border_gp = gpar(col = "black", lty = 1),row_names_side = "left")


htmp.list2 <- hm %v% htmp.colAnnot
plt.hm = draw(htmp.list2, 
     show_heatmap_legend = T, 
     ht_gap = unit(0.05, "cm"),)

saveplot(plt.hm, 'heatmap.olk.plsda.expression.with.loading', w = 10, h = 15)
```

# sparse PLS-DA on the olink data

Here we try to identify a core subset of proteins discriminating the two clusters.

## Model tuning to find the best number of features to include in the two PCs

Including 10 features at PC1 already reaches an error rate of 0.0017, and having more features or more PCs does not significantly increase prediction accuracy. So we just keep 10 features. We still keep 20 features on PC2 just for ordination plot. However, in later analysis with microbiome-proteome integration, we only use the 10 top features at PC1 for PLS and sPLS.

```{r splsda olink tune}
set.seed(666)
list.keepX <- c(1:9,  seq(10, 100, 10))
olk.plsda.tune <- tune.splsda(proteome.matrix.filtered,predictor, 
                              ncomp = 2, validation = 'Mfold', 
                              folds = 5, dist = 'max.dist', 
                              test.keepX = list.keepX, nrepeat = 100)

(f4 = plot(olk.plsda.tune, sd = T,optimal = F))
saveplot(f4,'olk.plsda.tune')

err = as.data.frame(olk.plsda.tune$error.rate)
create_dt(err,'Error rates')
```

## Tuned Olink sPLS-DA model

Including 10 features at PC1 already reaches an error rate of 0.0017, and having more features or more PCs does not significantly increase prediction accuracy. So we just keep 10 features. We still keep 20 features on PC2 just for ordination plot. However, in later analysis with microbiome-proteome integration, we only use the 10 top features at PC1 for PLS and sPLS.

```{r splsda tuned}
set.seed(666)

olk.splsda <- splsda(proteome.matrix.filtered,predictor,
                     ncomp = 2, keepX = c(10,20)) 

# ordination plot 
background.max <- background.predict(olk.splsda, 
                                     comp.predicted = 2,
                                     dist = 'max.dist') 
f.p = plotIndiv(olk.splsda, comp = c(1,2),
          ind.names = FALSE,
          ellipse = F, legend = TRUE,
          star = TRUE,
          title = 'PEA sPLS-DA',style = 'ggplot2',
          col = colorscheme_BAL_Protein_Cluster[1:2],
          background = background.max)
saveplot(f.p,'olk.splsda',w = 6)

# get features in PC1
olk.splsda.loading.pc1 <- olk.splsda$loadings$X %>%
  as.data.frame %>%
  dplyr::select(comp1) %>%
  dplyr::filter(comp1 != 0) %>%
  arrange(desc(abs(comp1))) %>%
  mutate(features = rownames(.))

# loadings

(f.l =ggbarplot(
  olk.splsda.loading.pc1,
  x = "features",
  y = "comp1",
  fill = "darkred",
  sort.val = "desc",
  sort.by.groups = FALSE,
  orientation = "h",         # Flip axes
  xlab = "Feature",
  ylab = "Loadings on PC1",
  palette = "jco"
) +
  theme_pubr() +
  theme(axis.text.y = element_text(size = 12)))
saveplot(f.l,'olk.splsda.loading.pc1')

proteome.selected = proteome.matrix.filtered[,rownames(olk.splsda.loading.pc1)]


```

# Identify biomarkers in the microbiome data

## PCA on the full microbiome data (after CLR + filtering low OTUs)

The PCA plot shows that there is no clear separation of the two clusters.

```{r 16s pca}
mb.pca <- pca(data.clr, ncomp = 2, scale = TRUE)
f1 = plotIndiv(mb.pca, group = predictor, ind.names = FALSE,
          legend = TRUE, 
          title = 'Microbiome PCA',
          style = 'ggplot2', star = TRUE,
          col = colorscheme_BAL_Protein_Cluster[1:2])
saveplot(f1, 'mb.pca',w=6)
```

# CCA and PLS of full microbiome and the selected proteins

CCA and PLS generate very similar results.
PLS regression of full microbiome data and the selected proteins show a generally trend in agreement in clustering between the two modalities. The two clusters can be separated on PC1, with some exceptions.

```{r 16s pls}

stopifnot(all(rownames(proteome.selected) == rownames(data.clr)))

mb.pls <- pls(data.clr,proteome.selected,ncomp = 2,mode = 'regression') 
print('PLS correlation coefficients: ')
cor(mb.pls$variates$X, mb.pls$variates$Y)
fp = plotIndiv(mb.pls, comp = c(1,2),
          ind.names = FALSE,
          #ellipse = TRUE, 
          legend = TRUE,
          star = TRUE,
          title = 'PLS with all OTUs and the selected proteins',style = 'ggplot2',
          group = predictor,
          col = colorscheme_BAL_Protein_Cluster[1:2],
          subtitle = c('Microbiome','PEA'))
saveplot(fp,'mb.olk.pls',w=10)
```

# sparse PLS 

Here we use sparse PLS to identify the core subset of OTUs that explains the most of the variance in the selected proteins while providing a good cluster separation pattern.

## Tuning the model to find the best number of OTUs to include in PCs 

```{r spls 16s tune,results='asis',fig.height=20,fig.width=20}
set.seed(666)
plot_list <- list()  # initialize an empty list
for (i in seq(150, 10, -10)){
  mb.spls <- spls(data.clr, proteome.selected,ncomp = 2,mode = 'regression',keepX = c(i,i))
  
  cor.res = cor.test(mb.spls$variates$X[,'comp1'], mb.spls$variates$Y[,'comp1'])
  cor.res.sel = paste0('coef=',round(cor.res$estimate[[1]],digits = 3))
  cat(cor.res.sel)
  
  cat('\n\n')
  p= plotIndiv(mb.spls, ind.names = FALSE, legend = TRUE, 
            title = paste0('nfeature = ',i, ', ',cor.res.sel),
            size.title = 20,
            group = predictor,centroid = T,
          style = 'ggplot2', star = TRUE,,
          col = colorscheme_BAL_Protein_Cluster[1:2],subtitle = c('Microbiome','PEA'),
          legend.position = 'top')$graph
  plot_list[[paste0("n", i)]] <- p
}

# Combine all plots (e.g., 3 columns)
combined_plot <- plot_grid(plotlist = plot_list, ncol = 5, align = "hv")

# Display or save
print(combined_plot)
saveplot(combined_plot,'spls.mb.olk.tuning.10-150features',w = 30,h = 15)
```

## sPLS with 50 OTUs (and 10 proteins)

```{r spls 16s}
set.seed(666)
mb.spls <- spls(data.clr,proteome.selected,ncomp = 2,mode = 'regression',keepX = c(50,50))

print('PLS correlation coefficients: ')
cor(mb.spls$variates$X, mb.spls$variates$Y)

f1 = plotIndiv(mb.spls, ind.names = FALSE, legend = TRUE, 
           title = 'sPLS with selected OTUs and proteins',group = predictor, centroid = T,
          style = 'ggplot2', star = TRUE,,
          col = colorscheme_BAL_Protein_Cluster[1:2],subtitle = c('Microbiome','PEA'))
saveplot(f1,'spls.mb.olk',w = 10)
```

```{r spls loadings,fig.height=24,fig.width=12}

otutaxa = tax_table(p.1r) %>% as.data.frame
mb.spls.loading.pc1 <- mb.spls$loadings$X %>%
  as.data.frame %>%
  dplyr::select(comp1) %>%
  dplyr::filter(comp1 != 0) %>%
  merge(.,otutaxa, by = 0, all.x = T) %>%
  mutate(abs_comp1 = abs(comp1),
         genus_otu = paste0(Row.names, ' (',Genus,')'),
         genus_otu = sub('Otu','OTU ',genus_otu)) %>%
  arrange(desc(abs_comp1)) 

mb.spls.olk.loading.pc1 <- mb.spls$loadings$Y %>%
  as.data.frame %>%
  dplyr::select(comp1) %>%
  dplyr::filter(comp1 != 0) %>%
  arrange(desc(abs(comp1))) %>%
  mutate(Row.names = rownames(.))

mb.spls.loading.pc1.combined = bind_rows(
  mb.spls.loading.pc1 %>%
    transmute(
      Features = genus_otu,
      comp1 = comp1,
      block = 'Microbiome'
    ) %>%
    arrange(desc(comp1)),
  mb.spls.olk.loading.pc1 %>%
    transmute(
      Features = Row.names,
      comp1 = comp1,
      block = 'PEA'
    )
) %>%
  mutate(Features = factor(Features,levels = Features),
         direction = case_when(comp1 > 0 ~ 'Positive',
                               comp1 < 0 ~ 'Negative'))
(f.l <- ggbarplot(
  mb.spls.loading.pc1.combined,
  x = "Features",
  y = "comp1",
  fill = "direction",
  sort.val = "none",
  sort.by.groups = FALSE,
  orientation = "h",          # horizontal bars (internally uses coord_flip)
  xlab = "Features",
  ylab = "Loadings on PC1",
) +
  theme_pubr() +
  theme(
    axis.title.x = element_text(size = 18),
    axis.title.y = element_text(size = 18),
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12),
    legend.position = "none"
  ) +
  scale_fill_manual(values = c("#2B5C8AFF","#AC4522FF")))


saveplot(f.l,'mb.spls.loading.pc1.combined',h=12,w=10)
```

## Heatmap (transformed count of each selected feature)

```{r heatmaps,fig.height=12,fig.width=20}
# PEA heatmap
## Main heatmap
mat = t(mb.spls$Y)
cat(paste0('NPX matrix min and max are: ', round(min(mat),digits = 2), ' and ', round(max(mat),digits = 2)))
## Add loading weightes on the right side as an one-column heatmap
loading_weight_vector <- mb.spls.olk.loading.pc1 %>%
  dplyr::select(comp1) %>%
  as.matrix()
loading_weight_vector <- loading_weight_vector[rownames(mat), , drop = FALSE]
col_fun_loading <- colorRamp2(c(-0.4, 0, 0.2), c("#2B5C8AFF", "#FFF6EDFF", "#AC4522FF"))

htmp.loading.annot.PEA <- rowAnnotation(`Loading weights` = loading_weight_vector,col = list(`Loading weights` = col_fun_loading),show_annotation_name = FALSE)


col_fun = colorRamp2(c(-2, 0, 2), c("orange", "#FFF6EDFF", "darkblue"))
hm = Heatmap(mat, name = 'PEA',
             column_title = "",
             show_row_names = T,show_row_dend = T,
             row_dend_width = unit(30,'mm'),
             column_dend_height = unit(20,'mm'),
             show_column_names = T, show_column_dend = T,
             right_annotation = htmp.loading.annot.PEA,
             cluster_columns = T,
             col = col_fun,
             border_gp = gpar(col = "black", lty = 1),
             clustering_method_rows = 'ward.D',
             row_names_max_width = unit(10,'cm')
)

# Microbiome heatmap
## main heatmap
mat = t(mb.spls$X[,mb.spls.loading.pc1$Row.names])
rownames(mat) <- mb.spls.loading.pc1$genus_otu[match(rownames(mat), mb.spls.loading.pc1$Row.names)]
cat(paste0('microbiome matrix min and max are: ', round(min(mat),digits = 2), ' and ', round(max(mat),digits = 2)))

## Add loading weight as an one-column heatmap to the right of the main heatmap
loading_weight_vector <- mb.spls.loading.pc1 %>%
  dplyr::select(genus_otu, comp1) %>%
  distinct() %>%
  tibble::column_to_rownames("genus_otu") %>%
  as.matrix()
loading_weight_vector <- loading_weight_vector[rownames(mat), , drop = FALSE]
htmp.loading.annot.microbiome <- rowAnnotation(`Loading weights` = loading_weight_vector,col = list(`Loading weights` = col_fun_loading),show_annotation_name = FALSE)


col_fun2 = colorRamp2(c(-2, 0, 5), c("orange", "#FFF6EDFF", "darkblue"))
hm2 = Heatmap(mat, name = 'Microbiome',
             column_title = "",
             show_row_names = T,show_row_dend = T,
             right_annotation = htmp.loading.annot.microbiome,
             row_dend_width = unit(30,'mm'),
             column_dend_height = unit(20,'mm'),
             show_column_names = T, show_column_dend = F,
             cluster_columns = F,
             col = col_fun2,
             border_gp = gpar(col = "black", lty = 1),
             clustering_method_rows = 'ward.D',
             row_names_max_width = unit(10,'cm')
)

# Add the outcome variable (olink cluster) as an one-row heatmap at the bottom
pred.vector <- matrix(predictor,
                      nrow = 1, 
                      dimnames = list("PEA Cluster", names(predictor)))
stopifnot(all(colnames(mat) == colnames(pred.vector)))

htmp.colAnnot <- Heatmap(pred.vector,
                         name = 'PEA Cluster',
                         show_row_names = TRUE,
                         show_row_dend = FALSE,
                         show_column_names = FALSE,
                         show_column_dend = FALSE,
                         col = colorscheme_BAL_Protein_Cluster,
                         height = unit(8, "mm"),
                         border_gp = gpar(col = "black", lty = 1),row_names_side = "left")

# Add the outcome variable (severity) as an one-row heatmap at the bottom
sev.vector <- phylo %>%
  sample_data() %>%
  as.matrix %>%
  as.data.frame %>%
  dplyr::select(disease_severity) %>% 
  as.matrix %>%
  t
sev.vector = sev.vector[, colnames(pred.vector),drop = FALSE]
stopifnot(all(colnames(sev.vector) == colnames(mat)))

htmp.colAnnot2 <- Heatmap(sev.vector,
                         name = 'Disease Severity',
                         show_row_names = TRUE,
                         show_row_dend = FALSE,
                         show_column_names = FALSE,
                         show_column_dend = FALSE,
                         col = colorscheme_severity,
                         height = unit(8, "mm"),
                         border_gp = gpar(col = "black", lty = 1),row_names_side = "left")


htmp.list2 <- hm %v% hm2 %v% htmp.colAnnot %v% htmp.colAnnot2
plt.hm = draw(htmp.list2, 
     show_heatmap_legend = T, 
     ht_gap = unit(0.05, "cm"),)

saveplot(plt.hm, 'heatmap.mb.olk.spls.expression.with.loading', w = 11, h = 10)


```

# Save session

```{r save session}
sessionInfo()
save.image(file.path(outdir,'Multiomics_analysis_sPLS_cleanCodes.Rdata'))

```
